---
title: "ATAC-seq Signal Normalization using edgeR"
author: "Quang Dang"
date: "2025-06-20"
output: html_document
---

This R Markdown document details the process of normalizing ATAC-seq data. The primary goal is to calculate scaling factors from raw read counts in peaks, which can then be used to generate normalized bigWig files for visualization.

### Load Required Libraries
First, we load the `purrr` library, which provides useful tools for functional programming, particularly for working with lists and functions.

```{r setup}
# purrr is used for its 'map' and 'reduce' functions, which simplify
# the process of reading and merging multiple files.
library(purrr)
# The tidyverse package (commented out) is a collection of R packages
# designed for data science, including dplyr and readr, which are used here.
# library(tidyverse)
```

---

### Part 1: Read and Merge FeatureCounts Data (Merged Replicates)
In this section, we read the output files from `featureCounts`. These files contain the number of reads from merged-replicate BAM files that fall within a set of consensus peak regions.

```{r read-merged-replicates}
# List all files in the specified directory that match the pattern
# "_readCountInPeaks.txt". This captures all the feature count outputs.
f_files <- list.files("/mnt/sdb/qdang/wt1_project/epi_mes_project/featurecounts",
                      pattern = "_readCountInPeaks.txt", full.names = TRUE)

# Define a function to read and process a single featureCounts file.
read_in_feature_counts <- function(file) {
    # Read the tab-separated file, treating the first line as a header
    # and ignoring any lines that start with a "#" comment character.
    cnt <- readr::read_tsv(file, col_names = TRUE, comment = "#")
    
    # The initial columns (Chr, Start, End, Strand, Length) describe the
    # peak regions. We only need the Geneid (peak ID) and the read counts,
    # so we remove these descriptive columns.
    cnt <- cnt %>% dplyr::select(-Chr, -Start, -End, -Strand, -Length)
    
    return(cnt)
}
    
# Use purrr::map to apply our reading function to every file in f_files.
# This creates a list of data frames, where each data frame contains the
# counts for one sample.
raw_counts <- map(f_files, read_in_feature_counts)

# Use purrr::reduce to iteratively merge all data frames in the list into
# a single data frame. The 'inner_join' function merges them based on the
# 'Geneid' column, ensuring that only peaks present in all samples are kept.
raw_counts_df <- purrr::reduce(raw_counts, inner_join)

# Clean up the column names by removing the suffix ".mLb.clN.sorted.bam"
# to make them shorter and more readable (e.g., "Sample1.bam" becomes "Sample1").
colnames(raw_counts_df) <- sub(".mLb.clN.sorted.bam", "", colnames(raw_counts_df))

# Save the final merged count matrix to a CSV file for future use or reference.
write.csv(raw_counts_df, file = "/mnt/sdb/qdang/wt1_project/epi_mes_project/featurecounts/countdata.csv")
```

---

### Part 2: Read and Merge FeatureCounts Data (Individual Replicates)
This chunk is similar to the first but processes count files from individual, non-merged replicates. This is often done to assess variability between replicates before merging.

```{r read-individual-replicates}
# List all featureCounts files for individual replicates.
f_files_indiv <- list.files("/mnt/sdb/qdang/wt1_project/epi_mes_project/featurecounts/individual_replicate",
                            pattern = "_readCountInPeaks.txt", full.names = TRUE)

# The same reading function from Part 1 is used here.
raw_counts_indiv <- map(f_files_indiv, read_in_feature_counts)
raw_counts_df_indiv <- purrr::reduce(raw_counts_indiv, inner_join)

# Clean up column names by removing the long path prefix.
colnames(raw_counts_df_indiv) <- sub("/mnt/sdb/qdang/wt1_project/epi_mes_project/", "", colnames(raw_counts_df_indiv))
# Further clean up by removing the BAM file suffix.
colnames(raw_counts_df_indiv) <- sub(".mRp.clN.sorted.bam", "", colnames(raw_counts_df_indiv))
```

---

### Part 3: Calculate Normalization Factors with edgeR
This is the core normalization step. We use the TMM (Trimmed Mean of M-values) method from the `edgeR` package to calculate normalization factors. TMM is robust against compositional bias, which can occur when a few genomic regions have extremely high signal in one sample compared to others.

```{r calculate-normalization-factors}
# Load the edgeR library, which is a standard tool for differential expression
# and normalization of count-based sequencing data.
library(edgeR)

# The edgeR functions require a numeric matrix with gene/peak IDs as row names.
# First, we set the 'Geneid' column as the row names of our data frame.
row.names(raw_counts_df) <- raw_counts_df$Geneid

# Convert all columns to numeric format. This is a crucial step because
# data read from files can sometimes be interpreted as characters or factors.
raw_counts_df2 <- data.frame(lapply(raw_counts_df, function(x) as.numeric(as.character(x))),
                           check.names = FALSE, row.names = rownames(raw_counts_df))

# Remove the original 'Geneid' column, which is now redundant because the
# peak IDs are stored as row names.
raw_counts_df2 <- subset(raw_counts_df2, select = -Geneid)

# Calculate TMM normalization factors. This function estimates the relative
# library sizes while accounting for compositional differences between samples.
NormFactor <- calcNormFactors(object = raw_counts_df2, method = "TMM")

# Get the total number of reads (library size) for each sample.
LibSize <- colSums(raw_counts_df2)

# Calculate the final size factors. These are what deepTools will use.
# The formula is (TMM Normalization Factor * Library Size) / 1,000,000.
# This produces a size factor relative to a library of one million reads (CPM scaling).
SizeFactors <- NormFactor * LibSize / 1000000

# For use with deepTools' bamCoverage, we need the reciprocal of the size factor.
# deepTools applies the scale factor by multiplication (Coverage = Reads * ScaleFactor),
# so we provide 1/SizeFactor to achieve normalization by division.
SizeFactors.Reciprocal <- 1 / SizeFactors

# Print the reciprocal factors. These are the values you will use with the
# --scaleFactor parameter in bamCoverage.
print(SizeFactors.Reciprocal)
```

---

### Part 4: Generate Normalized bigWig File
The final step is to use the calculated scale factor to generate a normalized coverage file in bigWig format. This is done outside of R using the `bamCoverage` tool from the deepTools suite.

```bash
# This command takes a BAM file and generates a bigWig file.
# --bam: The input alignment file.
# -o: The output normalized bigWig file.
# --binSize: The size of the genomic windows (bins) to calculate coverage over.
# --scaleFactor: The normalization factor. We use the reciprocal value calculated
#                in the R script above (e.g., 0.04026939 for this specific sample).
#                This effectively divides the coverage in each bin by the TMM-adjusted
#                library size, making it comparable to other samples.
# --numberOfProcessors: The number of CPU cores to use for faster processing.

bamCoverage --bam /mnt/sdb/qnap/qdang/andia_data/data_output_EPI_ATAC_E115E135E175/bowtie2/merged_library/E115_REP1.mLb.clN.sorted.bam \
 -o /mnt/sdb/qdang/atac/bigwig/E115_REP1.TMM.bw \
 --binSize 10 \
 --scaleFactor 0.04026939 \
 --numberOfProcessors 30